\myTitle

### 2025-03-25 tex/responses-to-referees/responses-to-reviewers-1.Rmd

Authors present on the development of a new set of markers, with many in close proximity to additional SNPs that can be used to assemble microhaplotypes. The marker set shows high discrimination power between the multiple ecotypes present with a large data set and fits well within the scope of the journal. Overall, the study is of excellent quality however, I have the following comments below.

_Thank you for your comments and suggestions.  We have addressed all of them, as
detailed below, and we are pleased with how this has improved the manuscript._

### Major Comments

* I’m confused by the amount of sequencing and a clear purpose of each sequencing run that should be clarified. There was ddRAD to identify novel polymorphisms/microhaplotypes, the second run using a using PCR-GBS type sequencing has an unclear purpose. Was this on the full population? Another “amplicon sequencing” run was performed to convert SNPtype polymorphisms, which sounds similar to the second sequencing run of ddRAD polymorphisms but uses different terminology. Was this part of the previous run? Was this run a few samples or the full population? I presume these were different runs but maybe they were the same. The GT-seq method, does this include all samples and the full set of markers to this point?
    - _We added to this section to make it more clear that the ddRAD one was to find gene regions, and the
    subsequent sequencing runs were made on amplicons from those gene regions, and we are also explicit about
    how many fish were used in those GT-seq optimization runs._

* Authors provide considerable detail on sample location in the methods but only provide the sample number in the introduction. I believe it would be useful to have summary sentences of Table 1 that include number sites for spring, fall, etc run, and the number of samples within each ecotype, total samples within the methods, etc.
    - _Thank you for this suggestion.  We added the following to the end of the first paragraph in the Population Sampling portion of the
Methods section:_
"Of the 1636 total fish samples from 17 different populations, 727 fish from 10 collections were of the fall run ecotype,
498 fish from 5 collections were spring run, and 111 and 300 fish were of the winter run and late-fall run ecotypes, 
respectively,  each represented by a single collection."

@@

* There is also use of informal language e.g. “amplified so well” or unnecessary words such as “then” that could be removed.
    - _Following this suggestion, we removed many occurrences of "then" and "as well as"_

* I commend the use of the distribution of log-likelihood ratio for relationship to show the power of haplotypes vs SNPs and would argue this could be elevated to a full figure, potentially in a condensed form, maybe less panels somehow?
    - _We appreciate the referee's feedback on this.  In response, we have created a
    new figure in the paper itself that examines the distribution of log-likelihood
    ratios only in the Feather River Hatchery.  This figure distinguishes between
    different log-likelihood ratios (PO/U, FS/U, and HS/U) and nicely shows the
    advantages of using microhaplotypes for relationship inference. We have also left
    the original figure in the supplement for any readers curious about the effect
    in different collections._

* If possible, I think it would also be good to show how much of an improvement the discrimination power improved from the old marker data. In theory, this data has already been generated for the full population if you strip out the novel SNPs of the current study.
    - _After long and careful consideration of this, we determined that it would not
    be possible to make an accurate comparison between the old and the new marker
    data in this regard.  Unfortunately, only 78 of the original 94 TaqMan assays
    were able to be converted into amplicon-sequenced markers, so we are not able
    to reproduce the old marker set with the same reference samples._

* It may also be worth discussing other studies on other organisms that either support or do not support the utilization and increased power achieved through haplotypes over SNPs.
    - _We have included four additional citation in the first paragraph of the discussion
    to place this work in the context of those other investigations._

@@

### Major Comments

* Lines 399-444: My biggest compliant is the paper states that it creates a “baseline” for GSI and PBT analysis but it is not always clear what this final baseline is and only at the end of the introduction does it state what this baseline contains (in terms of number of animals and number of loci). I feel this needs to also be stated in the abstract. 
    - _As per reviewer \#1's suggestion, the methods now includes textual comment on the
    number of samples from each collection. Additionally, we have added several sentences
    in the third paragraph of the introduction to define the terms reference data set and
    baseline, and we have also added a summary in the abstract
    on the total number of amplicons, individuals and populations in the reference data set._

* In addition, in the results section it is not easy for work out how many markers made into the final reference baseline, and I think this information should be included in a table that also shows how markers from each source of information. I’m assuming that this is what is being used in the subsequent analysis.
    - _Excellent suggestion.  This information now appears in the new Table 2 which shows the total number of
    amplicons and the total number of variant sites within each amplicon for each of several
    groups of markers that correspond to the different sources of information._

* Table 3: The values in second column (the average number of fish with observed genotypes) seem odd to me. I would have thought these values would be less than seventh column in Table 1 (column N). For example, CHLF has 379.6 in Table 3 > 261 in Table 1. Can the authors explain this? Furthermore, the proportion of individuals with observed genotypes is a more useful metric than the average number of non-missing genotypes, since there the number of fish between sampling sites is very different.
    - _Thank you very much for noting this.  We have now corrected this problem. The table was
    created from an old version of our metadata file that was missing entried from 42 fish.  These
    fish are now properly included in the metadata file and the figures in Table 1 have been
    regenerated. Also, we agree with the referee that the average number of non-missing genotypes
    is a less useful metric than one quantified as a proportion.  Therefore we have replaced
    the average number of non-missing genotypes with the total proportion of missing data in Table 1._

### 2025-03-25 tex/body-text-of-paper.tex
Done addressing comments (except Repo Readme)

Candidate loci with multiple single-nucleotide polymorphisms (SNPs) among multiple
populations were identified using reduced-representation genome sequencing of 2--4 Chinook salmon, each, from a number of
populations ranging from California to British
Columbia and the interior Columbia River (Table~\ref{tab:ascert-panel}).
Genomic DNA was extracted using Qiagen DNeasy 96 Blood and Tissue Kits on a BioRobot 3000
(Qiagen, Inc). Double-digest restriction-site associated DNA sequencing (ddRAD) with different size selections (400 bp and 500 bp
fragments) were used to produce a broad array of locus targets. Library preparation and sequencing methods followed those of
\citet{peterson2012double} with the modifications described in \citep{baetscher2018microhaplotypes}.
Sequencing was performed on a MiSeq (Illumina, Inc) using $2 \times 300$-cycle paired-end sequencing.

Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks were used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were then tested for PCR-amplification and consistency.  This step included
amplification in thousands of samples from a number of different collections of Chinook salmon for sampled
for several projects, with
subsequent sequencing on a
MiSeq sequencer using a $2\times 75$-cycle kit.
The sequences from each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

Basic data summaries were calculated for each population/collection
in the reference baseline
using simple operations in the R programming language, Version 4.3
\citep{rcore}. Summaries provided are:
\begin {itemize}
\item the average over loci of the number of fish
in which each locus was successfully genotyped, $\bar{N}$;
\item the average number of alleles per locus in each collection with all collections
downsampled to the smallest number of genotyped individuals at each
locus, $\bar{N}_\mathrm{A,ss}$;
\item the fraction of polymorphic
loci in each collection after subsampling to the smallest number of genotyped
fish, $\bar{P}_\mathrm{poly,ss}$;
\item the expected heterozygosity as the average over loci
of one minus the frequencies of homozygotes at each locus expected from the
estimated allele frequencies, $\bar{H}_\mathrm{exp}$; and
\item the observed heterozygosity as the average over
loci of the fraction of heterozygous genotypes, $\bar{H}_\mathrm{obs}$.
\end{itemize}
When subsampling each collection to the minimum number of genotyped
fish across all collections at a locus, individuals were subsampled without
replacement, taking the average of 1,000 different random subsamples.

@@

@@

A summary, which includes genomic location, consensus sequence, and primer sequences,
of all the amplicons used in the baseline is available
in Data Supplement~1.  An overview of genomic coordinates is found in Figure~\ref{fig:genome-locations}

@@

While there is not a currently available method \citep[like that of][]{anderson2006power}) for estimating error rates in
inference of parent-offspring trios with multiallelic marker data, given the substantial amount of statistical power for identifying
PO and FS pairs, which 
require much more statistical power, we hypothesize that PO-trio inference with these markers will be very accurate.

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations, and markers that have
long been in use for both GSI and PBT in California \citep{clemento2014evaluation}, by converting them
into so-called 'SNPlicons.' This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.

>>>>

Candidate loci with multiple single-nucleotide polymorphisms (SNPs) among multiple
populations were identified using reduced-representation genome sequencing of 2--4 Chinook salmon, each, from a number of
populations ranging from California to British
Columbia and the interior Columbia River (Table~\ref{tab:ascert-panel}).
Genomic DNA was extracted using Qiagen DNeasy 96 Blood and Tissue Kits on a BioRobot 3000
(Qiagen, Inc). Double-digest restriction-site associated DNA sequencing (ddRAD) with different size selections (400 bp and 500 bp
fragments) was used to produce a broad array of locus targets. Library preparation and sequencing methods followed those of
\citet{peterson2012double} with the modifications described in \citet{baetscher2018microhaplotypes}.
Sequencing was performed on a MiSeq (Illumina, Inc) using $2 \times 300$-cycle paired-end sequencing.
Stacks v1.45 \citep{catchen2013stacks} was used to analyze the ddRAD sequence data.  As there was not a well-assembled Chinook salmon genome
at the time, the ustacks module of Stacks were used to assemble reads {\em de novo} into RAD loci.  Initial filtering retained RAD loci with two or more SNPs present
in at least 10 samples. After this initial filtering, there were 3,931 RAD loci retained in the ddRAD data with 400 bp size selection and
4,794 retained in the data with 500 bp size selection.

The 8,725 discovered RAD loci were filtered down to a set of markers suitable for conversion
to amplicon sequencing assays. Gene regions were selected from the RAD loci that (1) contained multiple SNPs within a 100 bp window,
and (2) showed haplotype variation amongst the winter run and at least one other population.
A 100 bp window was used because the success of multiplex PCR and MiSeq sequencing is
diminished in amplicons larger than 100 bp. Primer3
software \citep{untergasser2012primer3} was used to design primers for 229 gene regions that met these criteria.
These 229 gene regions were then tested for PCR-amplification and consistency in a separate sample
of 288 Chinook salmon: 96 SRW fish and 48 each from the Entiat River (a Columbia River
tributary), TRH, FRHS, and CRLF\@.  Eight GT-seq amplifications were made and these were sequenced
on eight runs of a MiSeq sequencer
using a $2\times 75$-cycle kit to identify problematic loci and to optimize primer dilutions.
The sequences from each run at each locus were mapped to the {\em de novo} reference sequences (produced by Stacks) with
the Burrows-Wheeler aligner using the mem algorithm \citep{bwa-mem2009}. 
Subsequently, alignments were sorted with SAMtools \citep{li2009sequence} and provided to
FreeBayes \citep{garrison2012haplotype} for variant calling.
FreeBayes output was filtered to remove indels and retain variants with a Phred-scaled quality score of 30
or greater and a read depth of 10 or higher. Further filtering and analysis of the loci, including
assessment of Hardy-Weinberg equilibrium \citep{hardy1908mendelian} was conducted in MICROHAPLOT (Ng {\em et al.}, DOI: 10.5281/zenodo.820110).  Loci that produced more than two haplotypes within an individual, and those clearly violating Hardy-Weinberg equilibrium within populations were removed.

@@

Basic data summaries were calculated for each population/collection
in the reference baseline
using simple operations in the R programming language, Version 4.3
\citep{rcore}. Summaries provided are:
\begin {itemize}
\item the total fraction of missing genotypes, $M$;
\item the average number of alleles per locus in each collection with all collections
downsampled to the smallest number of genotyped individuals at each
locus, $\bar{N}_\mathrm{A,ss}$;
\item the fraction of polymorphic
loci in each collection after subsampling to the smallest number of genotyped
fish, $\bar{P}_\mathrm{poly,ss}$;
\item the expected heterozygosity as the average over loci
of one minus the frequencies of homozygotes at each locus expected from the
estimated allele frequencies, $\bar{H}_\mathrm{exp}$; and
\item the observed heterozygosity as the average over
loci of the fraction of heterozygous genotypes, $\bar{H}_\mathrm{obs}$.
\end{itemize}
When subsampling each collection to the minimum number of genotyped
fish across all collections at a locus, individuals were subsampled without
replacement, taking the average of 1,000 different random subsamples.

@@

The number of markers and variants obtained from each of our discovery efforts appears in
Table~\ref{tab:loci-summ}.  Further information, including genomic location, consensus sequence, and primer sequences,
of all the amplicons used in the reference data set is available
in Data Supplement~1.  A graphical overview of genomic coordinates is found in Figure~\ref{fig:genome-locations}.

@@

@@

While there is not a currently available method \citep[like that of][]{anderson2006power} for estimating error rates in
inference of parent-offspring trios with multiallelic marker data, given the substantial amount of statistical power for identifying
PO and FS pairs, which 
require much more statistical power, we hypothesize that PO-trio inference with these markers will be very accurate.

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations, and markers that have
long been in use for both GSI and PBT in California \citep{clemento2014evaluation}, by converting them
into so-called 'SNPlicons,' which are treated as multiallelic haplotypes in the presence
of additional variation surrounding the originally assayed SNP. Our work adds to a growing
chorus of recent studies finding that microhaplotype markers can provide additional
power over single SNPs for both GSI \citep{mckinney2017managing,mckinney2022new,hargrove2024efficient} and
relationship inference \citep{baetscher2018microhaplotypes,bootsma2020gt,delomas2024microhaplotypes}.
This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.
